Predicting the spread and persistence of genetically modified dominant sterile male mosquitoes

### 2021-07-29 inst/manuscript/Draft/bmc_article_revision_v1.tex

For this analysis we carefully elicited opinions on the mortality rate and dispersal distance of WT and DSM male mosquitoes from four independent experts who were not involved in the development of the DSM mosquitoes \cite{Hayes2015a}. Prior to the elicitation, experts were provided information on the genetic construct (originally incorporated into the G3 laboratory colony strain of \textit{An. gambiae}) and were told that the genetically modified strain would be repeatedly backcrossed with wild type mosquitoes captured in the vicinity of a biosecure laboratory in order to gradually replace the G3 genetic background with that of the locally originated strain. The location of the laboratory was provided to each expert but the species of local WT mosquitoes was not specified. For the mortality rate, experts were proposed to answer questions about either the probability of mortality, the probability of survival, or the life expectancy. All responses were converted to an estimate of the daily probability of survival ($p$) which was then converted to the mortality rate parameter ($\mu = -\log(p)$) used in the PDE model.

@@

Our posterior estimates of WT dispersal are conditional on a series of MRR experiments that were conducted prior to the elicitation of the dispersal prior distributions. These priors played no part in the experimental design, and the difference between the prior and posterior distributions depends on this design. This difference may reflect conservative prior estimates of mosquitoes' mean daily dispersal when in the vicinity of attractants (swarm locations) but it may also be an artefact of the MRR design that focussed efforts in and around village compounds to maximise the number of recaptures. As \cite{Epopa2017} notes, however, further intensive sampling studies outside or around villages will be useful to ensure that posterior estimates of dispersal are not blind to long range dispersal outcomes that are not witnessed because of the recapture site design.

>>>>

For this analysis we carefully elicited opinions on the mortality rate and dispersal distance of WT and DSM male mosquitoes from four independent recognised experts on malaria transmission by mosquitoes in Africa and genetic vector control methods; the independent experts were not involved in the development of the DSM mosquitoes \cite{Hayes2015a}. The experts were encouraged to draw on their own research experience and knowledge of the published literature while constructing their assessments.   Prior to the elicitation, experts were provided information on the genetic construct (originally incorporated into the G3 laboratory colony strain of \textit{An. gambiae}) and were told that the genetically modified strain would be repeatedly backcrossed with wild type mosquitoes captured in the vicinity of a biosecure laboratory in order to gradually replace the G3 genetic background with that of the locally originated strain. The location of the laboratory was provided to each expert but the species of local WT mosquitoes was not specified. For the mortality rate, experts were proposed to answer questions about either the probability of mortality, the probability of survival, or the life expectancy. All responses were converted to an estimate of the daily probability of survival ($p$) which was then converted to the mortality rate parameter ($\mu = -\log(p)$) used in the PDE model.

@@

Our posterior estimates of WT dispersal are conditional on a series of MRR experiments that were conducted prior to the elicitation of the dispersal prior distributions. These priors played no part in the experimental design, and the difference between the prior and posterior distributions depends on this design. This difference may reflect conservative prior estimates of mosquitoes' mean daily dispersal when in the vicinity of attractants (swarm locations) but it may also be an artefact of the MRR design that focussed efforts in and around village compounds to maximise the number of recaptures. As \cite{Epopa2017} notes, however, further intensive sampling studies outside or around villages will be useful to ensure that posterior estimates of dispersal are not blind to long range dispersal outcomes that are not witnessed because of the recapture site design. There is also the possibility of an impact of genetic modification on dispersal either through life history traits linked to vagility or perhaps host-seeking behavioural traits. If vagility is negatively impacted then dispersal could be less for DSM compared to wild-type, but if host-seeking ability was instead curtailed then dispersal could perhaps decrease for DSM relative to wild-type. As mentioned above, the analysis conducted here could be repeated to take into account the DSM MRR data that was in the planning stages at the time of the risk assessment.
